Where Do Events Come From?

### 2018-08-12 paper.tex
(>) Sugestoes Sztajnberg

\begin{abstract}
In reactive and event-based systems, execution is guided by an external
environment that generates inputs to the application and consumes outputs from
it.
Reactive languages provide dedicated syntax and semantics to deal with events
and greatly simplify the programming experience in this domain.
%
Nevertheless, the environment is typically prefabricated in a host language and
the very central concept of events is implemented externally to the reactive
language.
%
In this work, we propose an interrupt handler primitive for a reactive language
targeting embedded systems in order to take control of the whole event loop,
from input generation up to output effects.
%
We propose the new asynchronous primitive in the context of the synchronous
language \CEU and discuss how they synergize to avoid race conditions at
compile time, support lexically-scoped drivers, and provide automatic standby
for applications.
%and hot swapping of device drivers
\end{abstract}

@@

Reactive applications interact continuously and in real time with the external
world through hardware peripherals such as sensors and actuators (e.g.,
buttons, displays, timers, etc.).
%
These interactions are typically represented in software as input events
flowing from the peripherals to the application and as output events flowing
from the application to the peripherals.
%
%As illustrated in Figure~\ref{fig.env},
Peripherals can be abstracted in a
single component, \emph{the environment}, which connects with the application
through an event loop:
the application sits idle until the environment awakes it on
an input;
the application reacts and generates back outputs
which affect the environment;
the application becomes idle and the loop restarts.

@@

Reactive languages provide dedicated syntax and semantics to deal with events
and simplify the development of applications in this domain.
%
However, the environment is still typically implemented in a host language (e.g., C)
and controls the main event loop, invoking entry points into the reactive
language runtime on the occurrence of inputs, and also receiving output calls
from it.
%, both through a documented API.
%
\begin{comment}
As examples, Esterel~\cite{esterel.ieee91} relies on C for passing events
between the environment and the running program~\cite{esterel.book.compiling},
while Elm~\cite{frp.elm} uses the concept of \emph{ports}, which allows sending
out values to JavaScript as commands and listening for values as
subscriptions~\cite{frp.elm.ports}.
\end{comment}

@@

In this work, we propose interrupt service routines (ISRs) as an asynchronous
primitive for the synchronous reactive language \CEU~\cite{ceu.sensys13}.
ISRs empower reactive applications with their own device drivers which
self-generate inputs, bypassing the need for a host environment.
%
\CEU targets OS-less embedded architectures, such as Arduino-compatible
microcontrollers. %, in which applications run in the bare metal, without
%OS support.
%In this context, device drivers are typically libraries compiled together with
%the main program.
%
%The lack of an OS opens opportunities to explore a tighter integration between
%the application and its device drivers at the language level, resulting in an
%overall simplicity and tractability of the system.
%
%In particular, \CEU is already amenable to a simple static analysis to detect
%race conditions which we extended in this work to also encompass ISRs.
Applications in \CEU can share data buffers with drivers, to avoid unnecessary copying,
with some static guarantee that no data races will occur.
%
\CEU also provides a lexical finalization mechanism % that we adopt in drivers
which
safely disables drivers and turns off peripherals completely to save energy.
%For instance, we can delimit the scope of drivers similarly to lexically-scoped
%variables.
%
Finally, the synchronous semantics of \CEU enforces that applications react
to inputs in bounded time and remain in idle states susceptible to standby.
With the help of drivers and a power manager, the microcontroller
sleeps automatically at optimal sleeping modes after each input reaction.
%
%We implemented an extensible runtime that exposes hooks for the ISRs and
%power manager, which we validated in two microcontrollers: an
%\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
%
We implemented drivers for a variety of peripherals, such as GPIO, A/D
converter, USART, SPI, and an RF transceiver.
%
%The applications built on top of these drivers show optimal energy savings due
%to automatic standby.

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Asynchronous driver for GPIO which, on hardware interrupts, emits an
           input event into a queue.},
  label={lst.gpio.async},
  escapechar=\%,
]
// OUTPUT DRIVER

@@

The next example in Listing~\ref{lst.adc} illustrates the use of a
lexically-scoped driver for the A/D converter.
%
The application is a typical periodic sensor sampling loop, but which
designates an explicit \code{do-end} block (ln~20--24) to include the driver
(ln~21) and use it (ln~22--23).
%
The driver specifies a finalization code (ln~6--12) which executes
unconditionally whenever its enclosing block goes out of scope (ln~24).
The \code{finalize} completely disables all A/D functionality to save energy
(ln~8--10).
%
As discussed in the previous section, the finalizer would also execute if
aborted from a hypothetical \code{par/or} enclosing the driver.
%
Note that the driver interface (ln~3--4) is visible only inside the block
(ln~20--24) and thus cannot be inadvertently used outside it.

@@

Our runtime supports three compile-time configurations for \code{ceu\_pm\_sleep}
(ln~9):
%
The first configuration is a \emph{nop} which simply keeps the event loop
running all the time without ever sleeping.
This configuration is useful when introducing new platforms.
%
The second configuration chooses a (inefficient) sleep mode that keeps all
peripherals active.
Although this configuration is not the most energy efficient, at least, it
requires no extra assistance from the drivers.
%
The third sleep configuration keeps a bit vector of the active drivers during runtime
and chooses the most efficient mode, querying this vector every time
\code{ceu\_pm\_sleep} is called.

>>>>

\begin{abstract}
In reactive and event-based systems, execution is guided by an external
environment that generates inputs to the application and is affected by outputs
from it.
Reactive languages provide dedicated syntax and semantics to deal with events
and greatly simplify the programming experience in this domain.
%
Nevertheless, the environment is typically prefabricated in a host language and
the very central concept of events is implemented externally to the reactive
language.
%
In this work, we propose an interrupt handler primitive for a reactive language
that targets embedded systems in order to take control of the whole event loop,
from input generation up to output effects.
%
We propose the new asynchronous primitive in the context of the synchronous
language \CEU and discuss how they synergize to prevent runtime race conditions
at compile time, support lexically-scoped drivers, and provide automatic
standby for applications.
%and hot swapping of device drivers
\end{abstract}

@@

Reactive applications interact continuously and in real time with the external
world through hardware peripherals such as sensors and actuators (e.g.,
buttons, displays, timers, etc.).
%
These interactions are typically represented in software as input events
flowing from the peripherals to the application and as output events flowing
from the application to the peripherals.
%
%As illustrated in Figure~\ref{fig.env},
Peripherals can be abstracted in a
single component, \emph{the environment}, which connects with the application
through an event loop:
the application sits idle until the environment awakes it on
an input;
the application reacts and generates outputs back, which affect the environment;
the application becomes idle and the loop restarts.

@@

Reactive languages provide syntax and semantics specialized to deal with events
and simplify the development of applications in this domain.
%
However, the environment is still typically implemented in a host language (e.g., C)
and controls the main event loop, invoking entry points into the reactive
language runtime on the occurrence of inputs, and also receiving output calls
from it.
%, both through a documented API.
%
\begin{comment}
As examples, Esterel~\cite{esterel.ieee91} relies on C for passing events
between the environment and the running program~\cite{esterel.book.compiling},
while Elm~\cite{frp.elm} uses the concept of \emph{ports}, which allows sending
out values to JavaScript as commands and listening for values as
subscriptions~\cite{frp.elm.ports}.
\end{comment}

@@

In this work, we propose interrupt service routines (ISRs) as an asynchronous
primitive for the synchronous reactive language \CEU~\cite{ceu.sensys13}.
ISRs empower reactive applications with their own device drivers that
self-generate inputs, bypassing the need for a host environment.
%
\CEU targets OS-less embedded architectures, such as Arduino-compatible
microcontrollers. %, in which applications run in the bare metal, without
%OS support.
%In this context, device drivers are typically libraries compiled together with
%the main program.
%
%The lack of an OS opens opportunities to explore a tighter integration between
%the application and its device drivers at the language level, resulting in an
%overall simplicity and tractability of the system.
%
%In particular, \CEU is already amenable to a simple static analysis to detect
%race conditions which we extended in this work to also encompass ISRs.
Applications in \CEU can share data buffers with drivers, to avoid unnecessary copying,
with some static guarantee that no data races will occur.
%
\CEU also provides a lexical finalization mechanism % that we adopt in drivers
that safely disables drivers and turns off peripherals to save energy.
%For instance, we can delimit the scope of drivers similarly to lexically-scoped
%variables.
%
Finally, the synchronous semantics of \CEU enforces that applications react
to inputs in bounded time and remain in idle states susceptible to standby.
With the help of drivers and a power manager, the microcontroller
sleeps automatically at optimal sleeping modes after each input reaction.
%
%We implemented an extensible runtime that exposes hooks for the ISRs and
%power manager, which we validated in two microcontrollers: an
%\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
%
We implemented drivers for a variety of peripherals, such as GPIO, A/D
converter, USART, SPI, and an RF transceiver.
%
%The applications built on top of these drivers show optimal energy savings due
%to automatic standby.

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Asynchronous driver for GPIO that, on hardware interrupts, emits an
           input event into a queue.},
  label={lst.gpio.async},
  escapechar=\%,
]
// OUTPUT DRIVER

@@

The next example in Listing~\ref{lst.adc} illustrates the use of a
lexically-scoped driver for the A/D converter.
%
The application is a typical periodic sensor sampling loop, but which
designates an explicit \code{do-end} block (ln~20--24) to include the driver
(ln~21) and use it (ln~22--23).
%
The driver specifies a finalization code (ln~6--12) that executes
unconditionally whenever its enclosing block goes out of scope (ln~24).
The \code{finalize} completely disables all A/D functionality to save energy
(ln~8--10).
%
As discussed in the previous section, the finalizer would also execute if
aborted from a hypothetical \code{par/or} enclosing the driver.
%
Note that the driver interface (ln~3--4) is visible only inside the block
(ln~20--24) and thus cannot be inadvertently used outside it.

@@

Our runtime supports three compile-time configurations for \code{ceu\_pm\_sleep}
(ln~9):
%
The first configuration is a \emph{nop} that simply keeps the event loop
running all the time without ever sleeping.
This configuration is useful when introducing new platforms.
%
The second configuration chooses a (inefficient) sleep mode that keeps all
peripherals active.
Although this configuration is not the most energy efficient, at least, it
requires no extra assistance from the drivers.
%
The third sleep configuration keeps a bit vector of the active drivers during runtime
and chooses the most efficient mode, querying this vector every time
\code{ceu\_pm\_sleep} is called.
